---
title: "scfv_analysis"
author: "Mikolaj Kocikowski"
date: "2026-01-20"
output: html_document
---


```{r preparing-the-environment}

# GLOBAL PARAMETERS

# Required packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(circlize)
library(viridis)
library(corrplot)
library(reshape2)
library(ggalluvial)
library(ggpubr)
library(pheatmap)
library(tibble)
library(uwot)
library(svglite)
library(readr)


# Sample names to be analyzed (one per line in file) - must match colnames in df
sample_cols <- readLines("scfv_extraction_output/focus_libs.txt")

# Manual reordering to have naive library controls moved to the front (if present)
priority <- intersect(c("OG_LIB", "NEW_LIB"), sample_cols)
sample_cols <- c(priority, setdiff(sample_cols, priority))

# manual samples
# sample_cols <- c("COV_S1_R1", "COV_S1_R2", "COV_S1_R3", "COV_S1_R4")

# Grouping vector - same length as sample_cols - for grouping samples - not in use currently
# grouping_vector <- c("A", "A", "B", "B")

# Inferred linker - used for scfv space mapping - assumed constant for all scFvs
SCFV_LINKER_AA <- "EGKSSGASGESKVDDA"

# Create directories for output
OUTPUT_DIRS <- c(
  "r_output/alluvial",
  "r_output/barplots",
  "r_output/chords",
  "r_output/cooccurence",
  "r_output/diversity",
  "r_output/spaces",
  "r_output/tables"
)

for (d in OUTPUT_DIRS) {
  if (!dir.exists(d)) {
    dir.create(d, recursive = TRUE)
  }
}


```


```{r loading-raw-data}

# Load Seq2scFv output & library names
seq2scFv_output <- "scfv_extraction_output/in_frame_igBLAST_paired_delim_linker_scored_flags_counts.tsv"
libs <- "scfv_extraction_output/focus_libs.txt"

df <- read.table(
  seq2scFv_output,
  fill = TRUE,
  header = TRUE,
  sep = "\t",
  row.names = NULL
)

focus_libs <- read.table(libs)$V1

# Keep only libraries that are present in the loaded table.
missing_focus <- setdiff(focus_libs, colnames(df))
if (length(missing_focus) > 0) {
  warning("Dropping missing libraries from focus list: ", paste(missing_focus, collapse = ", "))
}
focus_libs <- setdiff(focus_libs, missing_focus)

missing_samples <- setdiff(sample_cols, colnames(df))
if (length(missing_samples) > 0) {
  warning("Dropping missing samples: ", paste(missing_samples, collapse = ", "))
}
sample_cols <- setdiff(sample_cols, missing_samples)

if (length(focus_libs) == 0 || length(sample_cols) == 0) {
  stop("No valid sample columns found in seq2scFv output.")
}

```


```{r filtering-the-data}

# Original (OG) filters with default settings - removing sequences with incorrect VH-VL order, bad linkers, NA in quality score 
filt_df_og <- df[!is.na(df$total_nt_qual), ] %>%
  filter(VH_IGH == 1,
         score_pcnt >= 75,
         mismatches <= 2,
         linker_overhang <= 10)

# some sequences with empty VH_cdr3_aa or VL_cdr3_aa slipped through the cracks; while potentially valid they can't be analyzed here
filt_df <- filt_df_og %>%
  filter(
    VH_cdr3_aa != "",
    VL_cdr3_aa != ""
  )

```


```{r collapsing-the-sequences}

# Collapse the sequences based on identical VH & VL (aa level)
VH_VL <- filt_df %>%
  group_by(VH_sequence_alignment_aa, VL_sequence_alignment_aa) %>%
  summarise(across(focus_libs[1]:total_nt_qual, sum), .groups = "drop")

# Collapse the sequences based on identical HCDR3 and LCDR3 (aa level)
CDR3s <- filt_df %>%
  group_by(VH_cdr3_aa, VL_cdr3_aa) %>%
  summarise(across(focus_libs[1]:total_nt_qual, sum), .groups = "drop")

# Collapse the sequences based on identical HCDR3 (aa level)
HCDR3 <- filt_df %>%
  group_by(VH_cdr3_aa) %>%
  summarise(across(focus_libs[1]:total_nt_qual, sum), .groups = "drop")

# Most entries appear only once even after collapsing; these are low-trust, since there is no enrichment process visible
VH_VL_nonone <- VH_VL %>%
  filter(total_nt_qual > 1)

CDR3s_nonone <- CDR3s %>%
  filter(total_nt_qual > 1)

HCDR3_nonone <- HCDR3 %>%
  filter(total_nt_qual > 1)

```

```{r saving-tables}

write.csv(filt_df, "r_output/tables/filt_df.csv", row.names = FALSE)
write.csv(VH_VL, "r_output/tables/VH_VL.csv", row.names = FALSE)
write.csv(CDR3s, "r_output/tables/CDR3s.csv", row.names = FALSE)
write.csv(HCDR3, "r_output/tables/HCDR3.csv", row.names = FALSE)
write.csv(VH_VL_nonone, "r_output/tables/VH_VL_nonone,.csv", row.names = FALSE)
write.csv(CDR3s_nonone, "r_output/tables/CDR3s_nonone.csv", row.names = FALSE)
write.csv(HCDR3_nonone, "r_output/tables/HCDR3_nonone.csv", row.names = FALSE)

```


```{r attrition-table-of-collapse}

# building a table showing how many sequences are left after certain filtering and collapsing steps in R (attrition table)

# store datasets and labels in a named list
datasets <- list(
  "Raw input data" = df,
  "Default-filtered raw data" = filt_df_og,
  "Non-empty HCDR3 and LCDR3" = filt_df,
  "Collapsed by VH and VL sequences" = VH_VL,
  "Collapsed by both CDR3s (VH and VL)" = CDR3s,
  "Collapsed by HCDR3" = HCDR3,
  "...VH & VL with total count > 1" = VH_VL_nonone,
  "...CDR3s with total count > 1" = CDR3s_nonone,
  "...HCDR3 with total count > 1" = HCDR3_nonone
)

# build attrition table
attrition_table_r <- data.frame(
  Step = names(datasets),
  Rows = sapply(datasets, nrow),
  row.names = NULL
)

# Export the table to CSV
write.csv(attrition_table_r,
          "r_output/tables/attrition_table_r.csv",
          row.names = FALSE)

attrition_table_r

```


```{r removing-dfs}

# Remove dfs that are no longer needed to free up RAM
rm(df)
rm(filt_df_og)
gc()
```


```{r merged-attrition-table}

# merging the R attrition table with the one generated from seq2scfv output files

# read TSV, skip first line (header)
attr_1 <- read.table(
  "scfv_extraction_output/attrition_table.tsv",
  sep = "\t",
  header = FALSE,
  stringsAsFactors = FALSE,
  skip = 1
)

# read CSV, skip first line (header) and second line (the same as last of the other table)
attr_2 <- read.csv(
  "r_output/tables/attrition_table_r.csv",
  header = FALSE,
  stringsAsFactors = FALSE,
  skip = 2
)

# combine tables
merged_attr <- rbind(attr_1, attr_2)

# set new column names
colnames(merged_attr) <- c("Stage", "Count")

# Export the table to CSV
write.csv(merged_attr,
          "r_output/tables/attrition_merged.csv",
          row.names = FALSE)

# print the result
merged_attr

```


```{r count-per-sample}

# Create summary data frame
summary_df <- data.frame(
  Sample = sample_cols,
  TotalCounts = colSums(VH_VL[, sample_cols]),
  UniqueSCFVs = sapply(VH_VL[, sample_cols], function(x)
    sum(x > 0))
)

summary_long <- summary_df %>%
  pivot_longer(
    cols = c(TotalCounts, UniqueSCFVs),
    names_to = "Metric",
    values_to = "Count"
  )


ggplot(summary_long, aes(
  x = factor(Sample, levels = sample_cols),
  y = Count,
  fill = Metric
)) +
  geom_col(
    position = position_dodge(width = 0.7),
    width = 0.6,
    color = "black"
  ) +
  scale_fill_manual(
    values = c(
      TotalCounts = "grey30",
      UniqueSCFVs = "grey75"
    ),
    labels = c("Total counts", "Unique scFvs")
  ) +
  labs(title = "Total counts vs Unique scFvs per sample",
       x = "",
       y = "Counts",
       fill = "") +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 90,
      hjust = 1, # modify if changing label angle
      vjust = 0.5 # modify if changing label angle
    ),
    axis.line = element_line(linewidth = 0.8),
    axis.ticks = element_line(linewidth = 0.8),
    legend.position = "top",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )


# Save the plot
ggsave(filename = "r_output/barplots/scfv_counts.png",
       plot = last_plot(),
       dpi = 300)
```
Effective Shannon Diversity summarizes how many clones are present and how evenly they are represented, making it a good single-number measure of library collapse under selection. In other words it tells me whether my library is broad or dominated by a few clones. Compared to simple ratios, it is also robust to sequencing depth differences, which makes it useful for comparing panning rounds.

```{r effectve-shannon-diversity}

effective_shannon <- numeric(length(sample_cols))

for (i in seq_along(sample_cols)) {
  counts <- VH_VL[[sample_cols[i]]]
  counts <- counts[counts > 0]
  
  p <- counts / sum(counts)
  
  # Shannon entropy
  shannon <- -sum(p * log(p))
  
  # Effective Shannon diversity
  effective_shannon[i] <- exp(shannon)
}

# Build tidy dataframe for plotting
div_df <- tibble(Sample = sample_cols, Effective_Shannon = effective_shannon)

ggplot(div_df, aes(
  x = factor(Sample, levels = sample_cols),
  y = Effective_Shannon)) +
  geom_col(width = 0.6,
           fill = "grey75",
           color = "black") +
  labs(title = "Effective Shannon diversity per sample", x = "", y = "Effective Shannon diversity") +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 90,
      hjust = 1, # modify if changing label angle
      vjust = 0.5 # modify if changing label angle
    ),
    axis.line = element_line(size = 0.8),
    axis.ticks = element_line(size = 0.8),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Save the plot
ggsave(filename = "r_output/diversity/Shannon_diversity_per_sample.png",
       plot = last_plot(),
       dpi = 300)
```


```{r diversity-metrics}

# prepare sample metadata. That could be easier if the samples always followed the same pattern, but controls don't
# doing this manually also lets us rename Antigen labels to something more telling, if needed

antigen_map <- tibble(
  Sample = c(
    "OG_LIB", "NEW_LIB",
    "BA1.1_R1", "BA1.1_R2", "BA1.1_R3", "BA1.1_R4",
    "BA5_R1", "BA5_R2", "BA5_R3", "BA5_R4",
    "COV_S1_R1", "COV_S1_R2", "COV_S1_R3", "COV_S1_R4",
    "Delta_R1", "Delta_R2", "Delta_R3", "Delta_R4",
    "SI_R1", "SI_R2", "SI_R3", "SI_R4",
    "envelope_z1_r1", "envelope_z1_r2", "envelope_z1_r3", "envelope_z1_r4",
    "trimer1_r1", "trimer1_r2", "trimer1_r3", "trimer1_r4",
    "trimer2_r1", "trimer2_r2", "trimer2_r3", "trimer2_r4",
    "CTL_S1_R1", "CTL_S1_R2", "CTL_S1_R3", "CTL_S1_R4"
  ),
  Antigen = c(
    "Naive_old", "Naive_fresh",
    rep("BA1.1", 4),
    rep("BA5", 4),
    rep("COV_S1", 4),
    rep("Delta", 4),
    rep("SI", 4),
    rep("envelope_z1", 4),
    rep("trimer1", 4),
    rep("trimer2", 4),
    rep("CTL_S1", 4)
  ),
  Round = c(
    1, 1,
    1,2,3,4,
    1,2,3,4,
    1,2,3,4,
    1,2,3,4,
    1,2,3,4,
    1,2,3,4,
    1,2,3,4,
    1,2,3,4,
    1,2,3,4
  ),
  Type = c(
    "Naive_old", "Naive_fresh",
    rep("Panning", 36)
  )
)


# compute diversity metrics per sample

diversity_df <- lapply(sample_cols, function(s) {
  counts <- VH_VL[[s]]
  counts <- counts[counts > 0]
  
  p <- counts / sum(counts)
  
  shannon <- -sum(p * log(p))
  effective_shannon <- exp(shannon)
  
  tibble(
    Sample = s,
    Shannon = shannon,
    Effective_Shannon = effective_shannon
  )
}) %>%
  bind_rows() %>%
  left_join(antigen_map, by = "Sample")

# Save data
write.csv(diversity_df, "r_output/tables/diversity_df.csv", row.names = FALSE)

# get antigen order to control facet order and start with controls
antigen_order <- antigen_map$Antigen %>% unique()

# Plot of effective diversity
ggplot(diversity_df, aes(x = factor(Round), # factor avoids weird x axis for controls due to having just 1 round
                         y = Effective_Shannon)) +  # I tried log, but it did not improve clarity
  # lines only for panning samples, skip it for controls
  geom_line(
    data = diversity_df %>% filter(Type == "Panning"),
    aes(group = Antigen),
    linewidth = 1
  ) +
  # points for all samples, including controls
  geom_point(size = 3) +
  # facet wrap for each target
  facet_wrap(~ factor(Antigen, levels = antigen_order), 
             scales = "free_x") + # fixed y makes facets comparable to each other
  labs(
    title = "Library diversity across panning rounds",
    x = "Panning round",
    y = "Effective Shannon diversity"
  ) +
  theme_classic(base_size = 14) +
  theme(
    panel.background = element_rect(fill = "grey95", color = NA),
    strip.background = element_rect(fill = "grey90", color = NA)
  )

# Save the plot
ggsave(filename = "r_output/diversity/Shannon_diversity_by_antigen.png",
       plot = last_plot(),
       dpi = 300)


##### Alternative, barplot based on previous, but with facets


ggplot(diversity_df, aes(x = Sample, y = Effective_Shannon)) +
  geom_col(width = 0.6, fill = "grey75", color = "black") +
  facet_wrap(~ factor(Antigen, levels = antigen_order), scales = "free_x", nrow = 1) +
  labs(
    title = "Effective Shannon diversity per sample",
    x = "",
    y = "Effective Shannon diversity"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    strip.text = element_blank(), # removing the facet desctiption boxes - to narrow to fit the antigen names nicely
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

# Save the plot
ggsave(filename = "r_output/diversity/Shannon_diversity_barstyle.png",
       plot = last_plot(),
       dpi = 300)
```

```{r distribution-counts-histogram}

# create a histogram of total_nt_qual, showing how frequently each count occurs across rows. Custom bins, otherwise value "1" towers over everything.

VH_VL_count_bins <- VH_VL %>%
  mutate(
    count_bin = case_when(
      total_nt_qual == 1 ~ "1",
      total_nt_qual <= 10 ~ "2-10",
      total_nt_qual <= 100 ~ "11-100",
      total_nt_qual <= 1000 ~ "101-1000",
      total_nt_qual > 1000 ~ "1001+"
    )
  ) %>%
  mutate(count_bin = factor(count_bin, levels = c("1", "2-10", "11-100", "101-1000", "1001+"))) # Ensure count_bin is an ordered factor, so that the bins show up in logical order

# Plot histogram with ordered bins

ggplot(VH_VL_count_bins, aes(x = count_bin)) +
  geom_bar(
    color = "black",
    fill = "grey75",
    alpha = 0.8,
    width = 0.6
  ) +
  scale_y_continuous(
    trans = 'log10',
    breaks = c(1, 10, 100, 1000, 10000),
    labels = scales::comma
  ) +
  labs(title = "Distribution of scFv Counts", x = "Total scFv Count Bins", y = "Frequency (counts - log scale)") +
  theme_classic(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(
      angle = 45,
      vjust = 1,
      hjust = 1
    ))

# Save the plot
ggsave(filename = "r_output/barplots/binned_counts_histogram.png",
       plot = last_plot(),
       dpi = 300)
```


Here we use a function to generate individual plots of public clonotypes, for any input datasets, and with a number in the tile.

```{r cooccurence-matrix}

plot_scfv_cooccurrence <- function(input_df,
                                   sample_cols,
                                   remove_diagonal = TRUE,
                                   df_name = NULL) {
  # Convert to logical: TRUE if scFv is present (>0)
  presence_mat <- input_df[, sample_cols] > 0
  
  # Compute co-occurrence matrix
  cooccur_mat <- t(presence_mat) %*% presence_mat
  
  # Name rows and columns
  rownames(cooccur_mat) <- colnames(cooccur_mat) <- sample_cols
  
  # OPTIONAL: remove diagonal
  if (remove_diagonal) {
    diag(cooccur_mat) <- NA
  }

  # Save it for stats
  write.csv(cooccur_mat,
          "r_output/tables/cooccur_mat.csv",
          row.names = FALSE)
  
  # Convert to long
  cooccur_long <- as.data.frame(cooccur_mat) %>%
    mutate(Sample1 = rownames(cooccur_mat)) %>%
    pivot_longer(-Sample1, names_to = "Sample2", values_to = "SharedSCFVs") %>%
    mutate(Label = ifelse(is.na(SharedSCFVs), "NA", SharedSCFVs))
  
  # Plot
  ggplot(cooccur_long, aes(x = Sample1, y = Sample2, fill = SharedSCFVs)) +
    geom_tile(color = "white") +
    # geom_text(aes(label = Label), size = 4) + # no space for printing values with high sample numbers
    scale_fill_gradient(
      low = "white",
      high = "red",
      space = "Lab",
      name = "Shared scFvs",
      na.value = "grey90"
    ) +
    labs(title = paste("Cooccurrence plot -", df_name)) +
    theme_minimal(base_size = 5) +
    theme(
      axis.text.x = element_text(
        angle = 45,
        vjust = 1,
        hjust = 1
      ),
      panel.border = element_rect(
        color = "black",
        fill = NA,
        size = 0.8
      )
    ) +
    coord_fixed()
}


##################### USAGE

# Queue of dataset names the "noone" versions won't differ other than on the diagonal
queue <- c("VH_VL", "CDR3s", "HCDR3")
# queue <- c("VH_VL")

for (name in queue) {
  p <- plot_scfv_cooccurrence(get(name), sample_cols, TRUE, name)
  print(p)
  ggsave(
    filename = paste0("r_output/cooccurence/Cooccurence_plot_", name, ".pdf"),
    plot = p,
    width = 4,
    height = 4
  )
}

```
Here we make a heatmap of public clonotypes, but as a single-use code for a faceted figure, no numbers in the tiles (not enough space).

```{r cooccurence-faceted}

# Define dfs to analyze and names to use for them on the plot
queue <- list("VH & VL" = VH_VL,
              "CDR3s"   = CDR3s,
              "HCDR3"   = HCDR3)

# Build combined long dataframe
cooccur_all <- bind_rows(lapply(names(queue), function(name) {
  input_df <- queue[[name]]
  presence_mat <- input_df[, sample_cols] > 0
  cooccur_mat <- t(presence_mat) %*% presence_mat
  rownames(cooccur_mat) <- colnames(cooccur_mat) <- sample_cols
  diag(cooccur_mat) <- NA  # remove diagonal values (somparison to the same sample)
  
  as.data.frame(cooccur_mat) %>%
    mutate(
      Sample1 = rownames(cooccur_mat),
      Dataset = factor(name, levels = names(queue)) # keep the order of facets as on the list
    ) %>%
    pivot_longer(-c(Sample1, Dataset),
                 names_to = "Sample2",
                 values_to = "SharedSCFVs") %>%
    mutate(Label = ifelse(is.na(SharedSCFVs), "NA", SharedSCFVs))
}))

# Prepare pseudocount column to enable using log in the plot coloring
cooccur_all <- cooccur_all %>%
  mutate(SharedSCFVs_plot = SharedSCFVs + 1)

# Plot
ggplot(cooccur_all, aes(x = Sample1, y = Sample2,  fill = SharedSCFVs_plot)) +
  geom_tile(color = "grey60", size = 0.25) + # set to white to change the look of tile border
  geom_text(aes(label = SharedSCFVs_plot), size = 0.75, color = "black") +  # add values on the plot
  scale_fill_gradient(
    low = "white",
    high = "red",
    na.value = "grey90",
    trans = "log10", # log is used to make the results more visible despite one scale and varying count levels between facets
    name = "Shared scFvs (+1, log10)" 
  ) +
  labs(x = NULL, y = NULL) +
  facet_wrap( ~ Dataset, nrow = 1) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 90,
      vjust = 1,
      hjust = 1,
      size = 4
    ),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_text(margin = margin(r = 5, b = 0), vjust = 0.75), # adjust vjust to get legend text on the level of legend bar
    axis.text.y = element_text(size = 4),
    strip.text = element_text(size = 12),
    panel.border = element_rect(
      colour = "black",
      fill = NA,
      linewidth = 1
    ),
    panel.grid = element_blank()
  ) +
  coord_fixed()

# Save the plot
ggsave(filename = "r_output/cooccurence/Cooccurence_faceted_numbers.png",
       plot = last_plot(),
       bg = "white",
       dpi = 300)

```

Building a heatmap with clustering for one chosen df: VH_VL


```{r pheatmap}

# Presence/absence partially normalizes for library sizes mismatch; this way the clustering is not focused on "expression" but types of scFvs

presence_mat <- VH_VL[, sample_cols] > 0
cooccur_mat <- t(presence_mat) %*% presence_mat

# Optional: hiding diagonal as its ofc high
# diag(cooccur_mat) <- 0

# prepating log data variant to visualize the whole range of abundances
cooccur_mat_log <- log10(cooccur_mat + 1)

pdf("r_output/cooccurence/Cooccurence_heatmap_clustered_VH_VL.pdf", width = 8, height = 8)

pheatmap(
  cooccur_mat_log, # log version of data
  cellheight = 10,
  cellwidth  = 10,
  display_numbers = FALSE, # numbers are confusing now that it log-ged data (and not just log fill like on the other plot)
  number_format = "%.0f", # no decimals
  number_color = "black",
  colorRampPalette(c("white", "red"))(20), # color = viridis::plasma(100)
  border_color = "grey70",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = TRUE,
  show_colnames = TRUE,
  treeheight_row = 50,
  fontsize = 6,
  fontsize_row = 6,
  fontsize_col = 6,
  main = "ScFv co-occurrence, presence/absence-based (log (Count + 1)), dataset: VH_VL)",
  clustering_distance_rows = "binary",
  clustering_distance_cols = "binary",
  clustering_method = "ward.D2" # cluster by shared presence patterns, not magnitude
)

dev.off()

```



```{r top-n-beauty-plots}
# Function to plot scFv dynamics across samples (smooth alluvial only)
plot_scfv_dynamics <- function(df,
                               samples,
                               ref_sample,
                               top_n = 10,
                               epsilon = 1e-6) {
  # Create synthetic (present in filtered df, but not in collapsed ones)
  df <- df %>% mutate(Clonotype = paste0("scFv_", row_number()))
  
  # Select top N clonotypes based on user-defined reference sample
  top_clones <- df %>%
    arrange(desc(.data[[ref_sample]])) %>%
    slice_head(n = top_n) %>%
    pull(Clonotype)
  
  # Pivot to long format
  plot_df_alluvial <- df %>%
    filter(Clonotype %in% top_clones) %>%
    select(Clonotype, all_of(samples)) %>%
    pivot_longer(
      cols = all_of(samples),
      names_to = "Sample",
      values_to = "Count",
      values_transform = list(Count = as.numeric)
    ) %>%
    mutate(
      Sample = factor(Sample, levels = samples),
      Count = ifelse(Count == 0, epsilon, Count)  # replace zeros with a small value, otherwise alluvial plotting will fail
    )

  # Build the ggalluvial plot
  ggplot(
    plot_df_alluvial,
    aes(
      x = Sample,
      y = Count / sum(Count[Sample == Sample]) * 100, # normalizing for library sizes
      alluvium = Clonotype,
      stratum = Clonotype,
      fill = Clonotype
    )
  ) +
    ggalluvial::geom_flow(width = 0.5, alpha = 0.8) +
    ggalluvial::geom_stratum(width = 0.5, color = "black") +
    scale_fill_viridis_d(option = "turbo") +
    scale_y_continuous(limits = c(0, 100), expand = c(0, 0)) +
    theme_pubr(legend = "right") +
    rotate_x_text(90) +
    theme(plot.title = element_text(hjust = 0.5)) +
    labs(
      x = "",
      y = "Relative abundance (%)",
      title = paste("Top", top_n, "scFvs from sample", ref_sample)
    )
}

####################### Usage

# Loop over each sample as reference and save
for (ref in sample_cols) {
  p <- plot_scfv_dynamics(VH_VL, sample_cols, ref_sample = ref, top_n = 10)
  print(p)
  ggsave(
    filename = paste0("r_output/alluvial/alluvial_top10_", ref, ".svg"),
    plot = p,
    width = 12,
    height = 6
  )
}

```


```{r}
# TEMPORARY: cleaning the environment to free up RAM, until I modularize

# Save an object to a file
saveRDS(filt_df, file = "filt_df.rds")

# MAnually RM, clean env

# Restore the object
filt_df <- readRDS(file = "filt_df.rds")
```


```{r umap-space}

# Why UMAP? Tsne was too slow with this data size already.

# Collapse sequences but keep the original sequence_id

VH_VL_collapsed <- filt_df %>%
  mutate(
    scfv_seq = paste0(
      VH_sequence_alignment_aa,
      SCFV_LINKER_AA,
      VL_sequence_alignment_aa
    ) # copied the infered linker sequence
  ) %>%
  group_by(scfv_seq) %>%
  summarize(sequence_id = first(sequence_id),
            across(all_of(sample_cols), \(x) sum(x, na.rm = TRUE)),
            .groups = "drop")


# Convert to long format (one row per scFv per sample)

scfv_long <- VH_VL_collapsed %>%
  pivot_longer(
    cols = all_of(sample_cols),
    names_to = "sample",
    values_to = "count"
  ) %>%
  filter(count > 0)


# Create feature matrix (k-mers)

# Why kmers instead of one-hot encoding? They capture similarity of motifs in varied sequences, while onehot requres aligned sequences of the same length.
# #  Kmer size 3; modify to ger more merged or fragmented clusters and regulate the computing load
k <- 3

kmerize <- function(seq, k = 3) {
  substring(seq, 1:(nchar(seq) - k + 1), k:(nchar(seq)))
}

# Unique scFv list
unique_seqs <- unique(scfv_long$scfv_seq)

# Build global k-mer vocabulary
all_kmers <- unique(unlist(lapply(unique_seqs, kmerize, k = k)))

# Create feature matrix
emb <- t(sapply(unique_seqs, function(s) {
  km <- kmerize(s, k)
  tab <- table(factor(km, levels = all_kmers))
  as.numeric(tab / sum(tab))   # normalize
}))

# save object - takes a long compute to get it!
saveRDS(emb, file = "emb_embedding.rds")

# UMAP embedding

n_neighbors <- min(15, nrow(emb) - 1)

um <- umap(emb, n_neighbors = n_neighbors, metric = "cosine")

# save object - takes a long compute to get it!
saveRDS(um, file = "um_umap.rds")

```

```{r preapare-umap-df}

# if needed
um <- readRDS(file = "um_umap.rds")

df_umap <- data.frame(scfv_seq = unique_seqs,
                      UMAP1 = um[, 1],
                      UMAP2 = um[, 2])

# merge

plot_df_umap <- scfv_long %>%
  left_join(df_umap, by = "scfv_seq")

# save the DF - it is computationally heavy to produce
  
write.csv(plot_df_umap, "r_output/tables/plot_df_umap.csv", row.names = FALSE)
  
```

```{r kde-plot}

# plot

ggplot(plot_df_umap, aes(UMAP1, UMAP2)) +
  stat_density_2d(
    aes(fill = after_stat(level)),
    geom = "polygon",
    alpha = 0.25,
    color = NA
  ) +
  geom_jitter( # or geom_point
    aes(color = sample),
    size = 0.5,
    alpha = 0.6
  ) +
  theme_classic(base_size = 14) +
  labs(title = "scFv sequence space (UMAP)") +
  theme(
    # removes meaningless labels
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.border = element_rect(
      color = "black",
      fill = NA,
      linewidth = 0.8
    ),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

# save the plot
ggsave(filename = "r_output/spaces/scFv_sequence_space_UMAP-KDE.png",
       plot = last_plot(),
       dpi = 300)

```


```{r kde-with-enrich}

# background KDE (all scFvs)
bg <- plot_df_umap %>%
  mutate(group = "all")

# enriched scFvs for a target (example) - select samples to show enrichment for
enriched <- plot_df_umap %>%
  filter(sample %in% c("COV_S1_R2", "COV_S1_R3", "COV_S1_R4", "COV_S1_R1"))

ggplot() +
  # full landscape
  stat_density_2d(
    data = bg,
    aes(x = UMAP1, y = UMAP2, fill = after_stat(level)),
    geom = "polygon",
    bins = 30,
    alpha = 0.4
  ) +
  
  # enriched contour outlines (no points); this will not plot but return warning for a sample with too few datapoints (probably some 50 min)
  stat_density_2d(
    data = enriched,
    aes(x = UMAP1, y = UMAP2, color = sample),
    geom = "contour",
    linewidth = 0.8
  ) +
  
  theme_classic() +
  labs(title = "Sequence space with enriched regions", color = "Selection round")

# Save the plot
ggsave(filename = "r_output/spaces/scFv_sequence_space_UMAP-KDE_contour.png",
       plot = last_plot(),
       dpi = 300)
```


```{r raster-per-sample}

# choose samples
# chosen_samples <- c("COV_S1_R2")
chosen_samples <- sample_cols

for (s in chosen_samples) {
  p <- ggplot() +
    geom_point(
      data = plot_df_umap,
      aes(UMAP1, UMAP2),
      color = "grey90",
      size = 0.2,
      alpha = 0.2
    ) +
    
    stat_density_2d(
      data = plot_df_umap %>% filter(sample == s),
      aes(UMAP1, UMAP2, fill = after_stat(density)),
      geom = "raster",
      contour = FALSE,
      n = 200
    ) +
    
    scale_fill_viridis_c() +
    coord_equal() +
    theme_classic() +
    ggtitle(s)
  
  print(p)
  
  ggsave(
    filename = paste0("r_output/spaces/UMAP_raster_", s, ".png"), # PDF and SVG not great here, because it is a raster image
    plot = p,
    width = 8,
    height = 6,
    dpi = 300
  )
}

```

```{r control-overlay-together}

# Choose your sample to plot
# chosen_samples <- c("COV_S1_R4")  # or loop through sample_cols
chosen_samples <- sample_cols

# Define all control rounds
control_samples <- c("CTL_S1_R1", "CTL_S1_R2", "CTL_S1_R3", "CTL_S1_R4",
                     "envelope_z1_r1", "envelope_z1_r2", "envelope_z1_r3", "envelope_z1_r4")

# Filter control data (all rounds combined)
control_data <- plot_df_umap %>% filter(sample %in% control_samples)

for (s in chosen_samples) {
  p <- ggplot() +
    # Background points
    geom_point(
      data = plot_df_umap,
      aes(UMAP1, UMAP2),
      color = "grey90",
      size = 0.2,
      alpha = 0.2
    ) +
    
    # Main sample density raster
    stat_density_2d(
      data = plot_df_umap %>% filter(sample == s),
      aes(UMAP1, UMAP2, fill = after_stat(density)),
      geom = "raster",
      contour = FALSE,
      n = 200
    ) +
    
    # All control contours combined (white)
    stat_density_2d(
      data = control_data,
      aes(UMAP1, UMAP2),
      color = "white",
      alpha = 0.6,
      bins = 10,
      linewidth = 0.4,
      contour = TRUE
    ) +
    
    scale_fill_viridis_c(option = "magma") +
    coord_equal() +
    theme_classic() +
    ggtitle(paste0(s, " (white = all control rounds combined)"))
  
  print(p)
  
#  ggsave(
#    filename = paste0("r_output/spaces/overlays/together/UMAP_raster_with_controls_together_", s, ".png"),
#    plot = p,
#    width = 8,
#    height = 6,
#    dpi = 300
#  )
}

```

```{r ontrols-overlay-separate}

# Choose your sample to plot
# chosen_samples <- c("COV_S1_R4")  # or loop through sample_cols
chosen_samples <- sample_cols

# Separate by control type
ctl_data <- plot_df_umap %>% filter(grepl("CTL_S1", sample))
envelope_data <- plot_df_umap %>% filter(grepl("envelope_z1", sample))

for (s in chosen_samples) {
  p <- ggplot() +
    geom_point(
      data = plot_df_umap,
      aes(UMAP1, UMAP2),
      color = "grey90",
      size = 0.2,
      alpha = 0.2
    ) +
    
    stat_density_2d(
      data = plot_df_umap %>% filter(sample == s),
      aes(UMAP1, UMAP2, fill = after_stat(density)),
      geom = "raster",
      contour = FALSE,
      n = 200
    ) +
    
    # CTL all rounds
    stat_density_2d(
      data = ctl_data,
      aes(UMAP1, UMAP2),
      color = "red",
      alpha = 0.7,
      bins = 5,
      linewidth = 0.7
    ) +
    
    # envelope_z1 all rounds
    stat_density_2d(
      data = envelope_data,
      aes(UMAP1, UMAP2),
      color = "cyan",
      alpha = 0.7,
      bins = 5,
      linewidth = 0.7
    ) +
    
    scale_fill_viridis_c(option = "magma") +
    coord_equal() +
    theme_classic() +
    ggtitle(paste0(s, " | Red = CTL (all rounds), Cyan = envelope_z1 (all rounds)"))
  
  print(p)
  
  ggsave(
    filename = paste0("r_output/spaces/overlays/separate/UMAP_raster_with_controls_separate_", s, ".png"),
    plot = p,
    width = 8,
    height = 6,
    dpi = 300
  )
}
```




```{r raster-facets}

# That's a good basic option with all samples

p <- ggplot(plot_df_umap, aes(UMAP1, UMAP2)) +
  stat_density_2d(
    aes(fill = after_stat(ndensity)),  # normalized density, showing each sample's structure well, like independent scales on single plots; also `density` does not work correctly
    geom = "raster",
    contour = FALSE,
    n = 200
  ) +
  scale_fill_viridis_c() +
  coord_equal() +
  theme_classic() +
  facet_wrap(~ sample, ncol = 4)

ggsave(
  filename = "r_output/spaces/UMAP_raster_faceted.png",
  plot = p,
  width = 8,
  height = 12,
  dpi = 300
)

print(p)
```



```{r raster-facets-reordered}

# It's a fancy version with samples chosen and ordered

# Define order: all samples except NEW_LIB and OG_LIB, then those two at the end
desired_order <- c(
  "BA1.1_R1", "BA1.1_R2", "BA1.1_R3", "BA1.1_R4",
  "BA5_R1", "BA5_R2", "BA5_R3", "BA5_R4",
  "COV_S1_R1", "COV_S1_R2", "COV_S1_R3", "COV_S1_R4",
  "CTL_S1_R1", "CTL_S1_R2", "CTL_S1_R3", "CTL_S1_R4",
  "Delta_R1", "Delta_R2", "Delta_R3", "Delta_R4",
  "SI_R1", "SI_R2", "SI_R3", "SI_R4",
  "envelope_z1_r1", "envelope_z1_r2", "envelope_z1_r3", "envelope_z1_r4",
  "trimer1_r1", "trimer1_r2", "trimer1_r3", "trimer1_r4",
#  "trimer2_r1", "trimer2_r2", "trimer2_r3", "trimer2_r4", # excluding trimer2, because only 1s sample shows correctly, and it confuses the plot order
  "OG_LIB", "NEW_LIB"  # Libraries at the end so they have a separate row
)

#plot_df_umap_ordered <- plot_df_umap %>%
#  mutate(sample = factor(sample, levels = desired_order))

plot_df_umap_ordered <- plot_df_umap %>%
  filter(sample %in% desired_order) %>%  # Only keep samples present in desired_order, meaning without the problematic trimer2; it is a duplicate anyway
  mutate(sample = factor(sample, levels = desired_order))

p <- ggplot(plot_df_umap_ordered, aes(UMAP1, UMAP2)) +
  stat_density_2d(
    aes(fill = after_stat(ndensity)),
    geom = "raster",
    contour = FALSE,
    n = 200
  ) +
  scale_fill_viridis_c(option = "viridis") +
  coord_equal() +
  theme_classic() +
  facet_wrap(~ sample, ncol = 4)

ggsave(
  filename = "r_output/spaces/UMAP_raster_faceted.png",
  plot = p,
  width = 8,
  height = 12,
  dpi = 300
)
print(p)
```

```{r faceted-control-overlay}
# Define all control rounds
control_samples <- c("CTL_S1_R1", "CTL_S1_R2", "CTL_S1_R3", "CTL_S1_R4",
                     "envelope_z1_r1", "envelope_z1_r2", "envelope_z1_r3", "envelope_z1_r4")

# Filter control data and REMOVE the sample column so it applies to all facets (avoids facetign by ~ sample)
control_data <- plot_df_umap %>% 
  filter(sample %in% control_samples) %>%
  select(-sample)

# Define order for facets
desired_order <- c(
  "CTL_S1_R1", "CTL_S1_R2", "CTL_S1_R3", "CTL_S1_R4", # controls first
  "envelope_z1_r1", "envelope_z1_r2", "envelope_z1_r3", "envelope_z1_r4", # controls first
  "BA1.1_R1", "BA1.1_R2", "BA1.1_R3", "BA1.1_R4",
  "BA5_R1", "BA5_R2", "BA5_R3", "BA5_R4",
  "COV_S1_R1", "COV_S1_R2", "COV_S1_R3", "COV_S1_R4",
  "Delta_R1", "Delta_R2", "Delta_R3", "Delta_R4",
  "SI_R1", "SI_R2", "SI_R3", "SI_R4",
  "trimer1_r1", "trimer1_r2", "trimer1_r3", "trimer1_r4",
  "OG_LIB", "NEW_LIB"
)

plot_df_umap_ordered <- plot_df_umap %>%
  filter(sample %in% desired_order) %>%
  mutate(sample = factor(sample, levels = desired_order))

p <- ggplot(plot_df_umap_ordered, aes(UMAP1, UMAP2)) +
  # Main density raster for each facet
  stat_density_2d(
    aes(fill = after_stat(ndensity)),
    geom = "raster",
    contour = FALSE,
    n = 200
  ) +
  
  # Control contours overlaid (same on all facets because no 'sample' column)
  stat_density_2d(
    data = control_data,
    color = "white",
    alpha = 0.6,
    bins = 5, # 2 minimum, 1 causes error
    linewidth = 0.3,
    contour = TRUE
  ) +
  
  scale_fill_viridis_c(option = "magma") +
  coord_equal() +
  theme_classic() +
  facet_wrap(~ sample, ncol = 4) +
  labs(caption = "White contours = combined negative controls (CTL + envelope_z1, all rounds)")

ggsave(
  filename = "r_output/spaces/umap_space_facets_controls_overlaid_together.png",
  plot = p,
  width = 8,
  height = 12,
  dpi = 300
)
print(p)

```

```{r saving-vertical}

# Define all control rounds
control_samples <- c("CTL_S1_R1", "CTL_S1_R2", "CTL_S1_R3", "CTL_S1_R4",
                     "envelope_z1_r1", "envelope_z1_r2", "envelope_z1_r3", "envelope_z1_r4")

# Filter control data and REMOVE the sample column so it applies to all facets
control_data <- plot_df_umap %>% 
  filter(sample %in% control_samples) %>%
  select(-sample)

# Create antigen and round columns
plot_df_umap_grid <- plot_df_umap %>%
  filter(!grepl("trimer2", sample)) %>%  # Remove the unwanted antigen, it only has 1 sample that plots correctly
  mutate(
    antigen = case_when(
      grepl("CTL_S1", sample) ~ "CTL_S1",
      grepl("envelope", sample) ~ "envelope_z1",
      grepl("BA1.1", sample) ~ "BA1.1",
      grepl("BA5", sample) ~ "BA5",
      grepl("COV_S1", sample) ~ "COV_S1",
      grepl("Delta", sample) ~ "Delta",
      grepl("^SI_", sample) ~ "SI",
      grepl("trimer1_", sample) ~ "trimer1", # careful so the regex does not catch ...r1 in the name as round
      sample == "OG_LIB" ~ "Libraries",
      sample == "NEW_LIB" ~ "Libraries",
      TRUE ~ sample
    ),
    round = case_when(
      grepl("_[Rr]1$", sample) ~ "R1",  # Must end with _R1 or _r1
      grepl("_[Rr]2$", sample) ~ "R2",  # Must end with _R2 or _r2
      grepl("_[Rr]3$", sample) ~ "R3",  # Must end with _R3 or _r3
      grepl("_[Rr]4$", sample) ~ "R4",  # Must end with _R4 or _r4
      sample == "OG_LIB" ~ "R1",
      sample == "NEW_LIB" ~ "R2",
      TRUE ~ "Other"
    )
  ) %>%
  mutate(
    antigen = factor(antigen, levels = c("CTL_S1", "envelope_z1", "BA1.1", "BA5", 
                                          "COV_S1", "Delta", "SI", "trimer1", "Libraries")),
    round = factor(round, levels = c("R1", "R2", "R3", "R4"))
  )

p <- ggplot(plot_df_umap_grid, aes(UMAP1, UMAP2)) +
  # Main density raster for each facet
  stat_density_2d(
    aes(fill = after_stat(ndensity)),
    geom = "raster",
    contour = FALSE,
    n = 200
  ) +
  
  # Control contours overlaid
  stat_density_2d(
    data = control_data,
    color = "white",
    alpha = 0.6,
    bins = 5,
    linewidth = 0.3,
    contour = TRUE
  ) +
  
  scale_fill_viridis_c(option = "magma") +
  coord_equal() +
  theme_classic() +
  facet_grid(antigen ~ round, switch = "y") +  # Rows = antigen, Cols = round
  theme(
    strip.placement = "outside",  # Put row labels outside
    strip.background = element_blank(),  # Remove grey background
    panel.spacing = unit(0, "lines")  # Reduce spacing between panels
  ) +
  labs(caption = "White contours = combined negative controls (CTL + envelope_z1, all rounds)")

ggsave(
  filename = "r_output/spaces/umap_space_facets_controls_overlaid_together_compressed.svg",
  plot = p,
  width = 10,
  height = 12,
  dpi = 300
)
print(p)

```




```{r chord-function}

# Generalizing chord diagram to a function so I can run it for each sample

plot_chord <- function(df, sample_col) {
  # count-weighted
  vh_vl_counts <- df %>%
    group_by(VH_v_call, VL_v_call) %>%
    summarise(count = sum(.data[[sample_col]]), .groups = "drop") %>%
    filter(count > 0)
  
  # presence / absence (unweighted)
  vh_vl_presence <- df %>%
    filter(.data[[sample_col]] > 0) %>%
    distinct(VH_v_call, VL_v_call) %>%
    mutate(count = 1)
  
  # SAFETY CHECKS - otherwise the function may stop and generate unreadable PDFs at samples with insufficient data
  
  if (nrow(vh_vl_counts) == 0 || nrow(vh_vl_presence) == 0) {
    message("Skipping ", sample_col, ": no VH–VL pairs")
    return(invisible(NULL))
  }
  
  if (length(unique(vh_vl_counts$VH_v_call)) < 1 ||
      length(unique(vh_vl_counts$VL_v_call)) < 1) {
    message("Skipping ", sample_col, ": not enough sectors for chord diagram")
    return(invisible(NULL))
  }
  
  # define colors for VH and VL families; okay to base it on counts and not presence dataset but use for both if both contain the same families
  vh_families <- unique(vh_vl_counts$VH_v_call)
  vl_families <- unique(vh_vl_counts$VL_v_call)
  
  # VH families: viridis palette
  vh_colors <- setNames(viridis(length(vh_families), option = "rocket"), vh_families)
  
  # VL families: magma palette
  vl_colors <- setNames(viridis(length(vl_families), option = "mako"), vl_families)
  
  # merge colors
  grid_colors <- c(vh_colors, vl_colors)
  
  # split plotting area
  par(
    mfrow = c(1, 2),
    mar = c(1, 1, 1, 1),
    oma = c(0, 0, 3, 0)
  ) # OMA needed for the main title
  
  # weighted plot (Count)
  chordDiagram(
    vh_vl_counts[, c("VH_v_call", "VL_v_call", "count")],
    grid.col = grid_colors,
    transparency = 0.25,
    annotationTrack = c("grid"),
    preAllocateTracks = 1
  )
  
  mtext("weighted (count-based)", side = 3, line = 0.5, cex = 1, outer = FALSE)
  
  circos.track(
    track.index = 1,
    panel.fun = function(x, y) {
      sector_name <- get.cell.meta.data("sector.index")
      xcenter <- get.cell.meta.data("xcenter")
      circos.text(
        x = xcenter,
        y = 0.9,
        labels = sector_name,
        facing = "clockwise",
        niceFacing = TRUE,
        adj = c(0.5, 0.5),
        cex = 0.5
      )
    },
    bg.border = NA
  )
  
  
  # unweighted plot (Presence-absence)
  chordDiagram(
    vh_vl_presence[, c("VH_v_call", "VL_v_call", "count")],
    grid.col = grid_colors,
    transparency = 0.25,
    annotationTrack = c("grid"),
    preAllocateTracks = 1
  )
  
  mtext("unweighted (presence-absence)", side = 3, line = 0.5, cex = 1, outer = FALSE)
  
  circos.track(
    track.index = 1,
    panel.fun = function(x, y) {
      sector_name <- get.cell.meta.data("sector.index")
      xcenter <- get.cell.meta.data("xcenter")
      circos.text(
        x = xcenter,
        y = 0.9,
        labels = sector_name,
        facing = "clockwise",
        niceFacing = TRUE,
        adj = c(0.5, 0.5),
        cex = 0.5
      )
    },
    bg.border = NA
  )
  
  
  # Title for the entire plot
  mtext(
    paste("VH-VL pairings -", sample_col),
    outer = TRUE,
    line = 1,
    cex = 1.2
  )
  
  # reset circlize state
  circos.clear()
}




########################## USAGE

# Use the function on any number of samples (count columns)
samples <- sample_cols

# overriding to test a few samples manually
# samples <- c("COV_S1_R1", "COV_S1_R2", "COV_S1_R3", "COV_S1_R4")

# Plot

for (s in samples) {
    # save the plot; it saves a file even if the function returned null due to security check (bad input)
    pdf(
    file = paste0("r_output/chords/chord_", s, ".pdf"), # careful with formats here
    width = 12,
    height = 6
  )
  plot_chord(filt_df, s)
  dev.off()
  plot_chord(filt_df, s) # display the plot; circlize does not return plots like normal ggplot to allow p <- function etc.
}


```
